---
title: "netZooR application with TB dataset"
author: "Tian Wang"
date: "2019-10-16"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{netZooR application with TB dataset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

netZooR is an R package which consists of  seven main algorithms and is able to construct, analyse and plot gene regulatory networks.

* **PANDA**(Passing Attributes between Networks for Data Assimilation) is a message-passing model to gene regulatory network reconstruction. It integrates multiple sources of biological data, including protein-protein interaction, gene expression, and transcription factor binding motifs information to reconstruct genome-wide, condition-specific regulatory networks.[[Glass et al. 2013]](http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0064832)

* **LIONESS**(Linear Interpolation to Obtain Network Estimates for Single Samples) is a method to estimate sample-specific regulatory networks by applying linear interpolation to the predictions made by existing aggregate network inference approaches.[[Kuijjer et al. 2019]]](https://www.sciencedirect.com/science/article/pii/S2589004219300872)

* **CONDOR** (COmplex Network Description Of Regulators) implements methods to cluster biapartite networks and estimatiing the contribution of each node to its community's modularity.[[Platig et al. 2016]](http://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1005033)

* **ALPACA**(ALtered Partitions Across Community Architectures) is a method to compare two genome-scale networks derived from different phenotypic states to identify condition-specific modules. [[Padi and Quackenbush 2018]](https://www.nature.com/articles/s41540-018-0052-5)

* **SAMBAR**(Subtyping Agglomerated Mutations By Annotation Relations) is a method to identify subtypes based on somatic mutation data.[[Kuijjer et al.]](https://www.nature.com/articles/s41416-018-0109-7).

* **MONSTER**(Modeling Network State Transitions from Expression and Regulatory data)[[Schlauch et al.]](https://doi.org/10.1186/s12918-017-0517-y): infers transcription factor which drivers of cell state conditions at the gene regulatory network level.

* **OTTER**(Optimization to Estimate Regulation) [publication in preparation]: models gene regulation estimation as a graph matrching problem


## Installation

### Prerequisites

Using this pacakage requires [**Python**](https://www.python.org/downloads/) (3.X) and some [Python libraries](#required-python-libraries), [**R**](https://cran.r-project.org/) (>= 3.3.3), and stable **Internet access**.

### Required Python libraries

How to install Python libraries depends varies from different platforms. More instructions could be find [here](https://packaging.python.org/tutorials/installing-packages/). 

The following Python libraries (or packages) are required by running PANDA and LIONESS algorithms:

The required Python packages are: [pandas](https://pandas.pydata.org/), [numpy](http://www.numpy.org/), [networkx](https://networkx.github.io/), [matplotlib.pyplot](https://matplotlib.org/api/pyplot_api.html).

### Installing
This package could be downloaded via `install_github()` function from `devtools` package.

```R
# install.packages("devtools") 
library(devtools)
# install netZooR pkg with vignettes, otherwise remove the "build_vignettes = TRUE" argument.
devtools::install_github("netZoo/netZooR", build_vignettes = TRUE)
library(viridisLite)#To visualize communities
```

## Data Resources

### Motif data
Here is some pre-prepared specie-sepcific PANDA-ready transcription factor binding motifs data stored in our AWS bucket https://s3.console.aws.amazon.com/s3/buckets/netzoo/netZooR/example_datasets/PANDA_ready_motif_prior/?region=us-east-2&tab=overview, which are derived from motif scan and motif info files located on https://sites.google.com/a/channing.harvard.edu/kimberlyglass/tools/resourcesby .

### PPI
This package includes a function `source.PPI` may source a Protein-Protein Interactions (PPI) througt STRING database given a list of proteins of interest. The [STRINGdb](http://www.bioconductor.org/packages/release/bioc/html/STRINGdb.html) is already loaded while loading netZooR.

```R
# TF is a data frame with single column filled with TFs of Mycobacterium tuberculosis H37Rv.
PPI <- source.PPI(TF, STRING.version="10", species.index=83332, score_threshold=0)
```

## Running the sample TB datasets


```{r}
library(netZooR)
```

Accessing the help pages for the  usage of core functions.
```R
?pandaPy
?createCondorObject
?pandaToCondorObject
?lionessPy
?alpaca
?pandaToAlpaca
?sambar
```

This package will invoke the Python in R environment through reticulate package.
Configure which version of Python to use if necessary, here in netZooR, Python 3.X is required. 
More details can be found [here](https://cran.r-project.org/web/packages/reticulate/vignettes/versions.html)
```R
#check your Python configuration and the specific version of Python in use currently
py_config()

# reset to Python 3.X if necessary, like below:
use_python("/usr/local/bin/python3")
```

The previous command is necessary to bind R to Python since we are calling PANDA from Python because netZooPy has an optimized implementation of PANDA. Check [this tutorial](http://netbooks.networkmedicine.org/user/marouenbg/notebooks/netZooR/panda_gtex_tutorial_server.ipynb) for an example using a pure R implementation of PANDA. 
Use example data sets within package to test this package.
Refer to four input datasets files: one TB expression dataset control group , one TB expression dataset treated, one transcription factor binding motifs dataset, and one protein-protein interaction datasets from either inst/extdat or AWS.

retrieve the file path of these files came with the netZooR package.
```{r}
# retrieve the file path of these files
treated_expression_file_path <- system.file("extdata", "expr4_matched.txt", package = "netZooR", mustWork = TRUE)
control_expression_file_path <- system.file("extdata", "expr10_matched.txt", package = "netZooR", mustWork = TRUE)
motif_file_path <- system.file("extdata", "chip_matched.txt", package = "netZooR", mustWork = TRUE)
ppi_file_path <- system.file("extdata", "ppi_matched.txt", package = "netZooR", mustWork = TRUE)
```

### PANDA algorithm

Assign the above file paths to flag `e`(refers to "expression dataset"), `m`(refers to "motif dataset"), and `ppi`(refers to "PPI" dataset), respectively. Then set option `rm_missing` to `TRUE` to run **PANDA** to generate an aggregate network without unmatched TF and genes.

Repeat with control group. 
```{r,message=FALSE,results="hide"}
treated_all_panda_result <- pandaPy(expr_file = treated_expression_file_path, motif_file = motif_file_path, ppi_file= ppi_file_path,modeProcess="legacy",  remove_missing = TRUE )
control_all_panda_result <- pandaPy(expr_file = control_expression_file_path,motif_file = motif_file_path, ppi_file= ppi_file_path,modeProcess="legacy",  remove_missing = TRUE )
```

Vector `treated_all_panda_result` and vector `control_all_panda_result` below are large lists with three elements: the entire PANDA network, indegree ("to" nodes) nodes and score, outdegree ("from" nodes) nodes and score. Use `$panda`,`$indegree` and `$outdegree` to access each list item resepctively.

Use `$panda`to access the entire PANDA network.
```{r}
treated_net <- treated_all_panda_result$panda
control_net <- control_all_panda_result$panda
```

### LIONESS Algorithm 
How to run LIONESS is mostly idential with method how to run PANDA in this package, unless the return values of `lionessPy()` is a data frame where first two columns represent TFs (regulators) and Genes (targets) while the rest columns represent each sample. each cell filled with estimated score calculated by LIONESS.

```{r,results="hide",message=FALSE}
# Run LIONESS algorithm for the first two samples
# removing start_sample and end_sample arguments to generate whole LIONESS network with all samples.
control_lioness_result <- lionessPy(expr_file = control_expression_file_path,motif_file = motif_file_path, ppi_file= ppi_file_path,modeProcess="legacy",  remove_missing = TRUE, start_sample=1, end_sample=2)
```

### CONDOR Algorithm and plotting
PANDA network can simply be converted into condor.object by `pandaToCondorObject(panda.net, threshold)`
Defaults option  `threshold` is the average of [median weight of non-prior edges] and [median weight of prior edges], all weights mentioned previous are transformationed with formula `w'=ln(e^w+1)` before calculating the median and average. But all the edges selected will remain the orginal weights calculated by PANDA.

```{r}
treated_condor_object <- pandaToCondorObject(treated_net, threshold = 0)
```

The communities structure can be plotted by igraph.
```{r}
library(viridisLite)
treated_condor_object <-condorCluster(treated_condor_object,project = FALSE)
treated_color_num <- max(treated_condor_object$red.memb$com)
treated_color <- viridis(treated_color_num, alpha = 1, begin = 0, end = 1, direction = 1, option = "D")
condorPlotCommunities(treated_condor_object, color_list=treated_color, point.size=0.04, xlab="Genes", ylab="TFs")
```

### ALPACA Algorithm

ALPACA community structure can also be generated from two PANDA network by `pandaToAlpaca`

```{r}
alpaca<- pandaToAlpaca(treated_net, control_net, NULL, verbose=FALSE)
```

## More tutorials

Browse with `browseVignettes("netZooR")`

## Information

```{r}
sessionInfo()
```

## Note
If there is an error like `Error in fetch(key) : lazy-load database.rdb' is corrupt` when accessing the help pages of functions in this package after being loaded. It's [a limitation of base R](https://github.com/r-lib/devtools/issues/1660) and has not been solved yet. Restart R session and re-load this package will help.

