---
title: "Quick-Starts"
author: "Tian Wang"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{QuickStart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# netZooR

netZooR is an R package which consists of five main algorithms and is able to construct, analyse and plot gene regulatory networks.

* **PANDA**(Passing Attributes between Networks for Data Assimilation) is a message-passing model to gene regulatory network reconstruction. It integrates multiple sources of biological data, including protein-protein interaction, gene expression, and transcription factor binding motifs information to reconstruct genome-wide, condition-specific regulatory networks.[[Glass et al. 2013]](http://journals.plos.org/plosone/article?id=10.1371/journal.pone.0064832)

* **LIONESS**(Linear Interpolation to Obtain Network Estimates for Single Samples) is a method to estimate sample-specific regulatory networks by applying linear interpolation to the predictions made by existing aggregate network inference approaches.[[LIONESS arxiv paper]](https://doi.org/10.1016/j.isci.2019.03.021)

* **CONDOR** (COmplex Network Description Of Regulators) implements methods to cluster biapartite networks and estimatiing the contribution of each node to its community's modularity.[[Platig et al. 2016]](http://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1005033)

* **ALPACA**(ALtered Partitions Across Community Architectures) is a method to compare two genome-scale networks derived from different phenotypic states to identify condition-specific modules. [[Padi and Quackenbush 2018]](https://www.nature.com/articles/s41540-018-0052-5)

* **SAMBAR**(Subtyping Agglomerated Mutations By Annotation Relations) is a method to identify subtypes based on somatic mutation data.[[Kuijjer et al.]](https://www.nature.com/articles/s41416-018-0109-7).



## Table of Contents
* [Getting Started](#getting-started) 
    + [Prerequisites](#prerequisites)
    + [Required Python libraries](#required-python-libraries)
    + [Installing](#installing)
* [Data Resources](#data-resources)
    + [Motif mapping data](#motif-mapping-data)
    + [PPI](#PPI)
* [Running the sample TB datasets](#running-the-sample-tb-datasets)
    + [PANDA algorithm](#panda-algorithm)
    + [PANDA Cytoscape Plotting](#panda-cytoscape-plotting)
    + [LIONESS Algorithm](#lioness-algorithm)
    + [CONDOR Algorithm and plotting](#condor-algorithm-and-plotting)
    + [ALPACA Algorithm](#alpaca-algorithm)
    + [SAMBAR Algorithm](#sambar-algorithm)
    + [MONSTER Algorithm](#monster-algorithm)
* [More tutorials](#more-tutorials)
* [Note](#note)
 


## Getting Started

### Prerequisites

Using this pacakage requires [**Python**](https://www.python.org/downloads/) (3.X) and some [Python libraries](#required-python-libraries), [**R**](https://cran.r-project.org/) (>= 3.3.3), and stable **Internet access**.

Some plotting functions will require the [**Cytoscape**](https://cytoscape.org/) installed.

### Required Python libraries

How to install Python libraries depends varies from different platforms. More instructions could be find [here](https://packaging.python.org/tutorials/installing-packages/). 

The following Python libraries (or packages) are required by running PANDA and LIONESS algorithms:

The required Python packages are: [pandas](https://pandas.pydata.org/), [numpy](http://www.numpy.org/), [networkx](https://networkx.github.io/), [matplotlib.pyplot](https://matplotlib.org/api/pyplot_api.html).

### Installing
This package could be downloaded via `install_github()` function from `devtools` package.

```{r}
# install.packages("devtools") 
library(devtools)
# install netZooR pkg with vignettes, otherwise remove the "build_vignettes = TRUE" argument.
devtools::install_github("netZoo/netZooR", build_vignettes = TRUE)
```

## Data Resources

### Motif data
Here is some [PANDA-ready transcription factor binding motifs data](https://sites.google.com/a/channing.harvard.edu/kimberlyglass/tools/resources) by species.

### PPI
This package includes a function `source.PPI` may source a Protein-Protein Interactions (PPI) througt STRING database given a list of proteins of interest. The [STRINGdb](http://www.bioconductor.org/packages/release/bioc/html/STRINGdb.html) is already loaded while loading netZooR.

```{r}
# TF is a data frame with single column filled with TFs of Mycobacterium tuberculosis H37Rv.
PPI <- source.PPI(TF, species.index=83332, score_threshold=0)
```

## Running the sample TB datasets

Use search() to check all loaded package currently.
```{r}
library(netZooR)
search()
```

Accessing the help pages for the  usage of core functions.
```{r}
?panda
?create.condor.object
?panda.to.condor.object
?lioness
?alpaca
?panda.to.alpaca
?sambar
```

This package will invoke the Python in R environment through reticulate package.
Configure which version of Python to use if necessary, here in netZooR, Python 3.X is required. 
More details can be found [here](https://cran.r-project.org/web/packages/reticulate/vignettes/versions.html)
```{r}
#check your Python configuration and the specific version of Python in use currently
py_config()

# reset to Python 3.X if necessary, like below:
# use_python("/usr/local/bin/python3")

```

Use example datasets within package to test this package.
Refer to four input datasets files: one TB expression dataset control group , one TB expression dataset treated, one transcription factor binding motifs dataset, and one protein-protein interaction datasets in inst/extdat.

```{r}
# retrieve the file path of these files
treated_expression_file_path <- system.file("extdata", "expr4.txt", package = "netZoo", mustWork = TRUE)
control_expression_file_path <- system.file("extdata", "expr10.txt", package = "netZoo", mustWork = TRUE)
motif_file_path <- system.file("extdata", "chip.txt", package = "netZoo", mustWork = TRUE)
ppi_file_path <- system.file("extdata", "ppi.txt", package = "netZoo", mustWork = TRUE)
```

### PANDA algorithm

Assign the above file paths to flag `e`(refers to "expression dataset"), `m`(refers to "motif dataset"), and `ppi`(refers to "PPI" dataset), respectively. Then set option `rm_missing` to `TRUE` to run **PANDA** to generate an aggregate network without unmatched TF and genes.

Repeat with control group. 
```{r}
treated_all_panda_result <- panda(e = treated_expression_file_path, m = motif_file_path, ppi = ppi_file_path, rm_missing = TRUE )
control_all_panda_result <- panda(e = control_expression_file_path, m = motif_file_path, ppi = ppi_file_path, rm_missing = TRUE )
```

Vector `treated_all_panda_result` and vector `control_all_panda_result` below are large lists with three elements: the entire PANDA network, indegree ("to" nodes) nodes and score, outdegree ("from" nodes) nodes and score. Use `$panda`,`$indegree` and `$outdegree` to access each list item resepctively.

Use `$panda`to access the entire PANDA network.
```{r}
treated_net <- treated_all_panda_result$panda
control_net <- control_all_panda_result$panda
```

### PANDA Cytoscape Plotting
Cytoscape is an interactivity network visualization tool highly recommanded to explore the PANDA network. Before using this function `plot.panda.in.cytoscape`, please install and launch Cytoscape (3.6.1 or greater) and keep it running whenever using.

```{r}
# select top 1000 edges in PANDA network by edge weight.
panda.net <- head(treated_net[order(control_net$force,decreasing = TRUE),], 1000)
 
# run this function to create a network in Cytoscape.
plot.panda.in.cytoscape(panda.net, network.name="PANDA")
```

### LIONESS Algorithm 
How to run LIONESS is mostly idential with method how to run PANDA in this package, unless the return values of `runLioness` is a data frame where first two columns represent TFs (regulators) and Genes (targets) while the rest columns represent each sample. each cell filled with estimated score calculated by LIONESS.

```{r}
# Run LIONESS algorithm for the first two samples
# removing start_sample and end_sample to generate whole LIONESS network with all samples.
control_lioness_result <- lioness(e = control_expression_file_path, m = motif_file_path, ppi = ppi_file_path, rm_missing = TRUE, start_sample=1, end_sample=2)
```

### CONDOR Algorithm and plotting
PANDA network can simply be converted into condor.object by `panda.to.condor.object(panda.net, threshold)`
Defaults option  `threshold` is the average of [median weight of non-prior edges] and [median weight of prior edges], all weights mentioned previous are transformationed with formula `w'=ln(e^w+1)` before calculating the median and average. But all the edges selected will remain the orginal weights calculated by PANDA.

```{r}
treated_condor_object <- panda.to.condor.object(treated_net, threshold = 0)
```

The communities structure can be plotted by igraph.
```{r}
treated_color_num <- max(treated_condor_object$red.memb$com)
treated_color <- viridis(treated_color_num, alpha = 1, begin = 0, end = 1, direction = 1, option = "D")
condor.plot.communities(treated_condor_object, color_list=treated_color, point.size=0.04, xlab="Target", ylab="Regulator")
```

### ALPACA Algorithm

ALPACA community structure can also be generated from two PANDA network by `panda.to.alpaca`

```{r}
alpaca<- panda.to.alpaca(treated_net, control_net, "./TB", verbose=TRUE)
```


### SAMBAR Algorithm

(working in progress)

### MONSTER Algorithm

(working in progress)

## More tutorials

Browse with `browseVignettes("netZooR")`

## Note
If there is an error like `Error in fetch(key) : lazy-load database.rdb' is corrupt` when accessing the help pages of functions in this package after being loaded. It's [a limitation of base R](https://github.com/r-lib/devtools/issues/1660) and has not been solved yet. Restart R session and re-load this package will help.

